---
title: "Week 4 chapter 5"
author: "Chi Hang(Philip) Cheung"
date: "2025-02-22"
output:
  html_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(fpp3)
```

5.1 Produce forecasts for the following series using whichever of
NAIVE(y), SNAIVE(y) or RW(y \~ drift()) is more appropriate in each
case:

Australian Population (global_economy) Only Naive and RW were used
because there is not seasonal pattern in the dataset.

```{r}
aus_pop <- global_economy %>% 
  filter(Country == 'Australia') %>% 
  select(Year, Population) %>% 
  mutate(Population = Population/1e6)

aus_fit<- aus_pop %>% model(
  Naive = NAIVE(Population),
  RW(Population ~ drift())
)
aus_pop_fc<- aus_fit %>% 
  forecast(h = '10 year')

aus_pop_fc %>% 
  autoplot(aus_pop, level = NULL)+
  labs(title='Aus population forecast', y='Population per millions')+
  guides(color = guide_legend(title = 'forecast models'))
```

Bricks (aus_production) SNAIVE is used due to seasonality pattern
observed.

```{r}
aus_brick<- aus_production %>% 
  select(Quarter, Bricks) %>% 
  filter(!is.na(Bricks))

aus_brick_fit<- aus_brick %>% 
  model(
    Snavie = SNAIVE(Bricks),
    )

aus_brick_fc<- aus_brick_fit %>% 
  forecast(h = 12)

aus_brick_fc %>% 
  autoplot(aus_brick, level = NULL)+
  labs(title = 'Aus Bricks production forecast')+
  guides(color = guide_legend(title = 'forecast'))
```

NSW Lambs (aus_livestock) SNAIVE and RW are used since there is a
seasonality pattern and a downward trend is observed.

```{r}
lamb<- aus_livestock %>% 
  filter(Animal == 'Lambs', State == 'New South Wales') %>% 
  mutate(Count = Count/1000)

lambs_fit<- lamb %>% 
  model(
    Snavie = SNAIVE(Count),
    RW(Count ~ drift())
  )

lambs_fc<- lambs_fit %>% 
  forecast(h = '5 year')

lambs_fc %>% 
  autoplot(lamb, level = NULL)+
  labs(title = 'Aus lambs count forecast', y='lambs count per thousands')+
  guides(color = guide_legend(title = 'forecast'))
```

Household wealth (hh_budget) Only Naive and RW are used due to lack of
seasonality from the data

```{r}
wealth<- hh_budget %>% 
  select(Year, Wealth)

wealth_fit<- wealth %>% 
  model(
    Naive = NAIVE(Wealth),
    RW(Wealth ~ drift())
  )

wealth_fc<- wealth_fit %>% 
  forecast(h = '5 year')

wealth_fc %>% 
  autoplot(wealth, level = NULL, scale='free_y')+
  labs(title = 'household wealth of 4 countries forecasts', y='USD')+
  guides(color = guide_legend(title = 'forecast'))
```

Australian takeaway food turnover SNAIVE and RW are used since the data
have an upward trend. Using NAIVE might be undermining the data trend.

```{r}
turnover<- aus_retail %>% 
  filter(Industry == 'Cafes, restaurants and takeaway food services', State== 'Australian Capital Territory') %>% 
  select(Month, Turnover)

turnover_fit<- turnover %>% 
  model(
    SNaive = SNAIVE(Turnover),
    RW(Turnover ~ drift())
  )

turnover_fc<- turnover_fit %>% 
  forecast(h = 24)

turnover_fc %>% 
  autoplot(turnover, level = NULL)+
  labs(title = 'Australian takeaway food turnover forecasts', y='turnover number')+
  guides(color = guide_legend(title = 'forecast'))
```

5.2 Use the Facebook stock price (data set gafa_stock) to do the
following:

a)  Produce a time plot of the series.

```{r}
#Since the trading days are irregular, a modification of the days are needed. Counted the days as the index.
fb<- gafa_stock %>% 
  filter(Symbol == 'FB') %>% 
  mutate(day = row_number()) %>% 
  select(day, Close) %>% 
  update_tsibble(index=day, regular = TRUE)

fb %>% autoplot(Close)+
  labs(y='closing stock price', title='Facebook stock time plot')
```

b)  Produce forecasts using the drift method and plot them.

```{r}
fb_fit<- fb %>% 
  model(RW(Close ~ drift()))

fb_fc<- fb_fit %>% 
  forecast(h = 720)

fb_fc %>% 
  autoplot(fb, level = NULL) +
  labs(title = 'fb stock random walk forecast', x='Trading days', y='Closing price')+
  guides(color = guide_legend(title = 'RW forecast'))
  
```

c)  Show that the forecasts are identical to extending the line drawn
    between the first and last observations.

```{r}
first_point<- fb %>% slice(1)
last_point<- fb %>% slice(n())

straight_line<- bind_rows(first_point, last_point)

fb %>%
  ggplot(aes(x = day, y = Close)) +
  geom_line(color = 'black') +  # Plot the historical data
  geom_line(data = straight_line, aes(x = c(1, nrow(fb)), y = c(first_point$Close, last_point$Close)), 
            color = 'red', linetype = "dashed") +  # Draw the straight line
  geom_line(data = fb_fc, aes(x=day, y=.mean), color='blue')+ #adding the forecast line
  labs(title = 'Facebook Stock Price with RW drift forecast method',
       x = 'Trading Days', y = 'Closing Price')
```

d)  Try using some of the other benchmark functions to forecast the same
    data set. Which do you think is best? Why?

    Ans: SNAIVE could not produce a forecast due to lack of seasonality.
    I believe the best benchmark tool to use for this data is the RW
    since it captures the growing trend of the fb stock in a long period
    of time. the Naive method is only applicable for companies that are
    stagnant.

```{r}
fb_fit2<- fb %>% 
  model(
    naive = NAIVE(Close),
    snaive = SNAIVE(Close)
  )
fb_fit2
fb_fc2<- fb_fit2 %>% forecast(h=720)

fb_fc2 %>% 
  autoplot(fb, level = NULL)+
  labs(title = 'Facebook Stock Price with Straight Line from First to Last Observation',
       x = 'Trading Days', y = 'Closing Price')
```

5.3 Apply a seasonal naïve method to the quarterly Australian beer
production data from 1992. Check if the residuals look like white noise,
and plot the forecasts. The following code will help. What do you
conclude?

Ans: The residuals look mostly center around the 0, indicating not much
of a sudden peak or trough for outliers. The ACF plot almost confirms
majority of the lags are in range with one except at lag 4. The residual
histogram also shows a relatively normal distribution. This is fair to
conclude that the residuals are most likely white-noise.

```{r}
recent_production <- aus_production |>
  filter(year(Quarter) >= 1992)
fit <- recent_production |> model(SNAIVE(Beer))
# Look at the residuals
fit |> gg_tsresiduals()
# Look a some forecasts
fit |> forecast() |> autoplot(recent_production)

```

5.4 Repeat the previous exercise using the Australian Exports series
from global_economy and the Bricks series from aus_production. Use
whichever of NAIVE() or SNAIVE() is more appropriate in each case.

Australian Exports series from global_economy:

Ans: NAIVE method is used since seasonality is lacking. The residuals
are normally distributed and the residual plots revolve around 0. The
Ljung_box test also shows p-value \> 0.05, indicating no autocorrelation
observed. Variations are due to white-noise only.

```{r}
aus_export<- global_economy %>% 
  filter(Country =='Australia') %>% 
  select(Year, Exports)

aus_fit<- aus_export %>% 
  model(naive=NAIVE(Exports))

aus_fit %>% gg_tsresiduals()
#check the p-value
augment(aus_fit) %>% features(.innov, ljung_box, lag=10)
#plotting the graph
aus_fit %>% forecast(h='5 year') %>% 
  autoplot(aus_export, level=NULL)+
  labs(title = 'Aus exports forecast')
```

Bricks series from aus_production Ans: SNAIVE is picked due to complex
seasonality is seen. The diagnostic plots show that the residuals are
left skewed from the histogram. ACF plot shows seasonal pattern of
autocorrelation, and the residual plot also shows outliers. The
residuals from this fitting are unlikely to be due to white noise. The
ljung_box test also confirms with a near or 0 p-value for the residual
autocorrelation.

```{r}
aus_brick<- aus_production %>% 
  select(Quarter, Bricks) %>% 
  filter(!is.na(Bricks))

aus_brick_fit<- aus_brick %>% 
  model(Snavie = SNAIVE(Bricks))
#diagnostic plots:
aus_brick_fit %>% gg_tsresiduals()
#ljung_box test:
augment(aus_brick_fit) %>% features(.innov, ljung_box, lag=10)
#Forecasted another 12 quarters
aus_brick_fc<- aus_brick_fit %>% 
  forecast(h = 12)

aus_brick_fc %>% 
  autoplot(aus_brick, level = NULL)+
  labs(title = 'Aus Bricks production forecast')+
  guides(color = guide_legend(title = 'forecast'))
```

5.7 For your retail time series (from Exercise 7 in Section 2.10): a)
Create a training dataset consisting of observations before 2011 using

```{r}
myseries <- aus_retail |>
  filter(`Series ID` == sample(aus_retail$`Series ID`,1))

myseries_train <- myseries |>
  filter(year(Month) < 2011)
```

b)  Check that your data have been split appropriately by producing the
    following plot.

```{r}
autoplot(myseries, Turnover, color='green') +
  autolayer(myseries_train, Turnover, color= 'red')
```

c)  Fit a seasonal naïve model using SNAIVE() applied to your training
    data (myseries_train).

```{r}
fit <- myseries_train |>
  model(snaive = SNAIVE(Turnover))
```

d)  Check the residuals. Do the residuals appear to be uncorrelated and
    normally distributed?

    Ans: The residuals look correlated and are not normally distributed.

```{r, warning=FALSE}
fit %>% 
  gg_tsresiduals()
```

e)  Produce forecasts for the test data

```{r}
fc_data<- myseries %>% 
  filter(year(Month)>2011)

fc<- fit %>% 
  forecast(new_data = anti_join(myseries, myseries_train))

fc %>% autoplot(myseries)
```

f)  Compare the accuracy of your forecasts against the actual values.

    Ans: The forecasted data is not very accurate when compared to the
    actual data. First, the plot shows almost no overlapping region
    between the observed data(black lines) and the forecasted
    data(purple). Second, all the forecast measurements are very big
    numbers when compared to the training data measurements. The SNAIVE
    method does not fit well with the model presented.

```{r}
fit %>%  accuracy() #training
fc %>% accuracy(myseries) #forecast value
```

g)  How sensitive are the accuracy measures to the amount of training
    data used?

Ans: The sensitivity of the accuracy measurement is relative important
to the amount of training data used. Typically accuracy improves as the
amount of training data is available for reference. In this case, the
forecasting method used is of benchmark purposes and the amount of
training data is slightly independent of the outcome due to how the
SNAIVE method takes the last seasonality patterns and replicate them. In
real forecasting models, the amount of training data will impact the
accuracy of the model more significantly.
