---
title: "Lab 4"
author: "Chi Hang (Philip) Cheung"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(RMySQL)
library(ggplot2)
```

### 1)read csv from db 1

```{r code-chunk-label}
host<- 'cunydata607sql.mysql.database.azure.com'
user<-Sys.getenv('user')
pw<- Sys.getenv('pw')
conn<- dbConnect(MySQL(),
                 user=user,
                 password=pw,
                 host=host,
                 dbname=user)
df<- dbGetQuery(conn, 'select * from delays')
print(df)
```

### 2)data tidying

dropping the index column

```{r dropping the index column}
df<- df %>%
  subset(select=c(2:8))
df
```

### 3)Transform the table into rows = observation and columns = variables only.

pivotlonger

```{r pivot_longer to tidy up the data}
df1<-pivot_longer(
    df,
    cols=c('Los Angeles', 'Phoenix', 'San Diego', 'San Francisco', 'Seattle'),
    names_to='Destination',
    values_to='Number of flights')
df1
```

```{r}
ggplot(df1, aes(x=Destination, y= `Number of flights`, fill=Status))+
  geom_bar(stat='identity', position='dodge') +
  facet_wrap(~Airline) +
  theme(
    axis.text = element_text(size = 6)
  )
```

## AM West has the highest On Time flights to Phoenix. This could also be that they have the most flights to Phoenix than Alaska Airline.

### 4)To compute the percentage of delay by location and airline:

```{r}
df2<- df1 %>%
  pivot_wider(
    names_from = Status,
    values_from =`Number of flights`
  )
df2<- df2 %>%
  rename('Number of On time flights'='On Time', 'Number of Delayed flights' = 'Delayed')
print(df2)
```

### 5)Adding a column of percentage of delay

```{r}
#Adding a column of percentage of delay
df2<- df2 %>%
  mutate(`Probability of a delay`=
           floor(100*`Number of Delayed flights`/(`Number of On time flights`+`Number of Delayed flights`)))
df2
```

```{r}
ggplot(df2, aes(x=Destination, y=`Probability of a delay`, fill = Airline)) +
  geom_bar(stat='identity', position = 'dodge') +
  scale_fill_manual(values=c('Alaska'='blue', `AM West`='red'))
```

### In conclusion, Alaska Airline has a lower overall probability of having a delay in their flights to the destinations mentioned above when compared to AM West. This is especially true for flights to San Francisco.
