---
title: "Lab 3"
author: "Chi Hang (Philip) Cheung"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(tibble)
```

### create an N1 table

```{r code-chunk-label}
N1<- tibble(
  EmployeeID = c(1, 1, 2, 3, 3),
  EmployeeName = c("John", "John", "Jane", "Alice", "Alice"),
  Department = c("Sales", "Sales", "Marketing", "HR", "HR"),
  Project = c("Project A", "Project B", "Project C", "Project D", "Project E"),
  HoursWorked = c(10, 15, 20, 30, 25)
)

N1
```

**This table is in N1 form because all the data are atomic and the rows and columns are not repeating.**

## transform N1 table to N2

**Since the table is in N1 form, we can transform it into N2, eliminating partial dependencies on non-key columns.**

```{r}
N2EmpID<- N1 %>%
  subset(select=c(EmployeeID, EmployeeName, Department)) %>%
  distinct()
N2EmpID
```

**To further eliminate partial dependency, we will introduce a new column called ProjectID. This makes the Project column fully dependent on the primary key, which is ProjectID rather than the EmployeeID in the original table.**

```{r}
N2D_Pro<- N1 %>%
  mutate(ProjectID= c('AA01', 'AA02', 'AA03', 'AA04', 'AA05')) %>%
    subset(select=c(ProjectID, Project)) %>%
      distinct()
print(N2D_Pro)
```

```{r}
N2Hour<- N1 %>%
  mutate(ProjectID= c('AA01', 'AA02', 'AA03', 'AA04', 'AA05')) %>%
    subset(select=c(EmployeeID, ProjectID, HoursWorked)) %>%
      distinct()
print(N2Hour)
```

**These three tables happen to also satisfy N3F since each non-key attributes are fully dependent on the primary keys in each table and no transitive dependency is observed. The non-key attribute "HoursWorked" is fully dependent on the composite keys "EmployeeID" and "ProjectID".**

## Data filtering. 173 majors

```{r Filtering 173 College majors}
major<- read.csv('https://raw.githubusercontent.com/stormwhale/607-assignment/refs/heads/main/majors-list.csv')
major<-data.frame(major)


name<- major %>%
  filter(str_detect(Major, regex('DATA|STATISTIC', ignore_case = TRUE)))
name
```

## 3.  Describe, in words, what these expressions will match:

(.)\\1\\1 **The correct expression should be (.)\\\\1\\\\1 and it will match three of the same characters captured in the first group. (e.g. aaa, ccc, bbb)**

"(.)(.)\\2\\1" **\<- This will capture 4 characters that are reverse of one another. (e.g saas)**

(..)\\1 **\<- The correct expression should be "(..)\\\\1" and it will match anything that is repeating the first two characters. e.g. abab**

"(.).\\1.\\1" **\<- this captures any first letter with any letter that follows and returns the first captured letter with another letter finally followed by the first letter. E.g "cbcnc"**

"(.)(.)(.).\*\\3\\2\\1" **\<- captures three characters followed by a string that could be anything and any length as long as the reverse of the first three characters are followed. E.g. xyz123abc123zyx**

## 4.  Construct regular expressions to match words that:

Start and end with the same character.

```{r}
df<- 'aaafffddf'
print(str_view(df, "(.)\\1"))
```

Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)

```{r}
df2<- c("church", "success", "letter", "banana", "bookkeeper")
str_view(df2, "(.)(.).*\\1\\2")
```

Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)

```{r}
df3<- 'Tennessee'
str_view(df3, "(.).*\\1.*\\1")
```
