---
title: "Lab 5 Project1"
author: "Chi Hang(Philip) Cheung"
date: "2024-10-03"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)

#path to dataset
url<- "https://raw.githubusercontent.com/stormwhale/607-assignment/refs/heads/main/tournamentinfo.txt"
```

## To read the txt file:

```{r}
df<- read.delim(url, sep="|", header=FALSE,skip=4)
head(df)
               
```

## To transform the data into the correct dataframe:

```{r}
rows_to_remove<- seq(from= 0, to = nrow(df), by=3) #removing extra separator lines
clean_df<- df[-rows_to_remove, ] #anything before the "," indicates the row
head(clean_df)

#filter out every second row and merge two data frames together into one:
second_row<- seq(2, nrow(clean_df), by =2)
second_row_df<-clean_df[second_row, ]
head(data.frame(second_row_df))

first_row_df<-clean_df[-second_row, ]

head(first_row_df)

#to combine the two dataframes:
com_df<- cbind(first_row_df, second_row_df)
head(com_df)
```

## remove extra columns and rename existing columns:

```{r}
com_df<- com_df %>%
  subset(select=c(1:10, 12:13))

colnames(com_df)<- c("PairID", "Name", "Points", "Round1", "Round2", "Round3", "Round4", "Round5", "Round6", "Round7", "State", "USCF_ID")
head(com_df)
```

## To make the values atomic in each cell
```{r}
com_df<- com_df %>%
  separate_wider_delim('USCF_ID',
                       delim="/",
                       names= c('USCF_ID',
                                'Pre_rating'))
```

## To extract ONLY the pre rating from the Pre_rating column:

```{r}
com_df$Pre_rating<- trimws(str_sub(com_df$Pre_rating, 4,8))
print(com_df)
```

## To extract only PairID from the rounds:

```{r}
com_df[, 4:10]<- com_df[, 4:10] %>%
  lapply(function(x) trimws(gsub("[^0-9]", "", x)))

#to convert columns to be numeric:
com_df[,c(1, 3:10, 12:13)]<- lapply(com_df[,c(1, 3:10, 12:13)], as.numeric)
print(com_df)
```

# Alternative way but less accurate:

com_df[, 4:10]\<- com_df[, 4:10] %\>% lapply(function(x) trimws(str_sub(x, 3,5))) view(com_df)

## To create a function in searching the PairID with opponent pre-rating:

```{r}
fetch<- function(df) {
  lapply(df, function(x) {
  x<- x[!is.na(x)] #to remove NA from matching
  
  if (length(x)==0) {
    return(NA)
  }
  P_rating<- com_df$Pre_rating[com_df$PairID == x]
  
  if (length(P_rating) > 0) {
    return(P_rating)
  } else {
    return(NA)}
  })}
#transformation to make each round PairID in exchange for the pre_rating
com_df[, 4:10]<- lapply(com_df[, 4:10], fetch)
view(com_df)
# need to convert the Rounds back to numeric again:
com_df[, 4:10]<- lapply(com_df[, 4:10], as.numeric)
summary(com_df)
```

## To swap the column positions:

```{r}
com_df<- com_df %>%
  mutate(Avg_op_rating = floor(rowMeans(com_df[1:64, 4:10], na.rm = TRUE)))
print(com_df)
```

## To reaarange the columns for CSV export:

```{r}
final<- com_df %>%
  subset(select=c(Name, State, Points, Pre_rating, Avg_op_rating))
print(final)
```

## Save to CSV:

```{r}
write.csv(final, file='Project_Chess.csv', row.names = FALSE)
```
