---
title: "607 Lab 2 SQL DB connection"
author: "Chi Hang (Philip) Cheung"
date: "`r Sys.Date()`"
output:  pdf_document
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
```
### Hiding login information:
```{r info hide}
library(rstudioapi)
username<-showPrompt(title='username', message='Enter username:', default="")
pw<-askForPassword(prompt="Enter password:")
```

###Forming DB connection

```{r code-chunk-label}
library(RMySQL)
mydb <- dbConnect(MySQL(), user=username, password=pw, dbname=username, host="cunydata607sql.mysql.database.azure.com")
dbListTables(mydb)
```

### Load the movies table into R dataframe
```{r}
#write SQL command:
df<- dbGetQuery(mydb, "select * from movies")
df
```
### to view the dataframe summary
```{r}
df<-df %>%
  mutate('averageScore'= floor(rowMeans(select(., where(is.numeric)), na.rm=TRUE)))
df 
```
```{r plot}
library(ggplot2)
ggplot(df, aes(x = Movies, y = averageScore, color=Movies, fill=Movies, legend=FALSE)) +
  geom_bar(stat='identity') +
  labs(title = 'Average ratings from 6 reviews of the 6 popular movies', x='Movie names')+
  theme(legend.position = 'none')
```

```{r close connection}
#dbDisconnect()
```

## Part 4: Missing data strategy
There are few approaches that i would take depending on the nature of the miss values.
First, the proportion of the missing data is more than 50% of the columns total data, i might consider excluding the column from my analysis. Huge amount of missing data sometimes can be more misleading than providing meaningful insights
Second, if the missing data is small (>20%) and is numerical, i will consider to use simple imputation or mean imputation to fill in the missing data with the column values' mean.
If the missing data are categorical, it will require addition analysis or even predictive model to try filling in the missing value