---
title: "607 Lab 6 Project2"
author: "Chi Hang(Philip) Cheung, Inna Yedzinovich"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

```{r load-packages}
options(repos = c(CRAN = "https://cran.r-project.org"))
library(tidyr)
suppressPackageStartupMessages(library(dplyr))
library(readr)
library(ggplot2)
library(tidyverse)
```

## Dataset 1: SAT scores in 2010

```{r}
url1<- "https://raw.githubusercontent.com/stormwhale/data-mines/refs/heads/main/SAT__College_Board__2010_School_Level_Results.csv"
df2<- read.csv(url1)
head(df2)
```
### To tidy up and transform the data:
```{r}
df2_tidy<- df2 %>%
  pivot_longer(cols = c('Critical.Reading.Mean', 'Mathematics.Mean', 'Writing.Mean'),
               names_to = 'Test_subjects',
               values_to = 'Average_scores',
               names_pattern ='(.*)\\.Mean')
df2_tidy<-rename(df2_tidy,c('Number_of_test_takers'='Number.of.Test.Takers', 'School_name'='School.Name'))

#drop NA schools
df2_tidy<- df2_tidy %>%
  subset(!is.na(Number_of_test_takers))

#check if any NA still exist:
any(sum(is.na(df2_tidy)))
head(df2_tidy)
```

### Analyzing the data

To get the top 10 schools ranked by total SAT scores:

```{r}
top_total<-df2_tidy %>%
  group_by(School_name) %>%
  summarise(total_SAT_score= sum(Average_scores)) %>%
  slice_max(total_SAT_score, n= 10)

print(top_total)
```

### Top 10 schools SAT SCOREs break down by each test subject

```{r}
top_sub <- df2_tidy %>%
  filter(School_name %in% top_total$School_name)
#To get the mean for each test subject from the top 10 schools:
top_sub_mean<- top_sub %>%
  group_by(Test_subjects) %>%
  summarise(Mean_score = mean(Average_scores))
print(top_sub_mean) 
```

```{r}
ggplot(top_sub, aes(x=reorder(School_name, Average_scores), y= Average_scores, fill=Test_subjects))+
  geom_bar(stat='identity') +
  coord_flip() +
  labs(title = 'Top 10 High Schools SAT scores by Subject', y='', x ='Average SAT Scores')+
  theme(legend.position = 'bottom')
```

### To look at how far ahead the top 10 schools from the mean scores of other schools:

```{r}
#To get the mean and standard deviation from all HS that took the SAT:
tot_stat<- df2_tidy %>%
  group_by(Test_subjects) %>%
  summarise(Mean_score= mean(Average_scores), SD=sd(Average_scores), '1+SD'=Mean_score+SD, '2+SD'=Mean_score+SD*2)

tot_stat_mean<- tot_stat$Mean_score
tot_stat_2sd<- tot_stat$`2+SD`

ggplot() +
  geom_histogram(data= df2_tidy, aes(x=Average_scores), fill='grey', binwidth=30)+
  geom_histogram(data= top_sub, aes(x=Average_scores), fill='blue', binwidth=30)+
    facet_wrap(~Test_subjects)+
  labs(title = 'Top 10 high schools mean SAT score distribution by test subject',
        y='',
        x='Average Scores') +
  geom_vline(data= tot_stat, aes(xintercept= tot_stat_mean, color ='Mean'), linetype='dotted', linewidth = 0.9)+
  geom_vline(data= tot_stat, aes(xintercept= tot_stat_2sd, color ='2+SD'), linetype='dashed', linewidth=0.9)+
  scale_color_manual(values= c('Mean'='red', '2+SD'='black'))
```

## Conclusion: Top 10 High Schools are above 2 standard Deviation from the mean scores

### Correlation between \# of test takers and average SAT scores

```{r}
ggplot(df2_tidy, aes(x= Average_scores, y=Number_of_test_takers))+
  geom_point() +
  facet_wrap(~Test_subjects) +
  geom_smooth(method='lm', se=FALSE) +
  labs(x= 'Average SAT Score')
```

There is a positive correlation between number of test takers and average SAT scores. The higher number of people from a school participating in the SAT, the average score is generally higher than the mean value. This is observed in all three test subjects.

## Dataset 2: Unity Data - MTA Daily Ridership

Data Overview The dataset contains daily ridership and traffic data for
various transportation modes in New York City during March 2020. The
columns include:

-   Date: The date of the record.
-   Subways: Total estimated ridership and percentage of comparable
    pre-pandemic day.
-   Buses: Total estimated ridership and percentage of comparable
    pre-pandemic day.
-   LIRR (Long Island Rail Road): Total estimated ridership and
    percentage of comparable pre-pandemic day.
-   Metro-North: Total estimated ridership and percentage of comparable
    pre-pandemic day.
-   Access-A-Ride: Total scheduled trips and percentage of comparable
    pre-pandemic day.
-   Bridges and Tunnels: Total traffic and percentage of comparable
    pre-pandemic day.
-   Staten Island Railway: Total estimated ridership and percentage of
    comparable pre-pandemic day.

Initial Analysis:\
- There is a noticeable decline in ridership across all transportation
modes as the month progresses. - The percentage of ridership compared to
pre-pandemic levels shows a significant decline. - Access-A-Ride: This
services maintained higher percentages of pre-pandemic levels compared
to other modes, indicating continued demand for these services despite
the pandemic - Traffic through bridges and tunnels also decreased but
not as drastically as public transportation ridership. This could
suggest a shift towards private vehicle usege during the pandemic.

```{r MAT-Riders-DataSet}
url <- "https://raw.githubusercontent.com/Yedzinovich/Data-607/main/MTA_Daily_Ridership_Data.csv"

mta_data <- read_csv(url)
mta_data <- mta_data %>% mutate(Date = as.Date(Date, format = "%m/%d/%y"))

mta_data_long <- mta_data %>% pivot_longer(cols = -Date, names_to = "Metric", values_to = "Value")
head(mta_data_long)

mta_data_long <- mta_data_long %>% separate(Metric, into = c("Transport_Mode", "Metric_Type"), sep = ": ")
head(mta_data_long)
```

Now that we have the data in a long format, we can extract more
comprehensive insights from it. Long format can help us to perform a
variety of analyses that are more flexible and insightful compared to
the original wide format.

\*\*\*What do know: - March 11, 2020, marks the start of the federal
COVID-19 PHE declaration. - May 11, 2023, marks the end of the federal
COVID-19 PHE declaration.
Source:<https://archive.cdc.gov/www_cdc_gov/coronavirus/2019-ncov/your-health/end-of-phe.html#>:\~:text=The%20federal%20COVID%2D19%20PHE,and%20testing%2C%20will%20remain%20available.

```{r MAT-Riders-DataSet-Analysis}
# Analysis #1
avg_ridership <- mta_data_long %>%
  filter(Metric_Type == "Total Estimated Ridership") %>%
  group_by(Transport_Mode) %>%
  summarize(Average_Ridership = mean(Value, na.rm = TRUE))

print(avg_ridership)

# Analysis #2
ggplot(mta_data_long %>% filter(Metric_Type == "Total Estimated Ridership"), aes(x = Date, y = Value, color = Transport_Mode)) +
  geom_point() +
  labs(title = "Public Transportation Ridership Trends Over Time", x = "Date", y = "Total Estimated Ridership")

# Analysis #3
percentage_mta_data <- mta_data_long %>%
  filter(grepl("% of Comparable Pre-Pandemic Day", Metric_Type))

ggplot(percentage_mta_data, aes(x = Date, y = Value, color = Transport_Mode)) +
  geom_point() +
  labs(title = "Percentage of Pre-Pandemic Levels Over Time",
       x = "Date",
       y = "Percentage of Pre-Pandemic Levels",
       color = "Transport Mode")

# Analysis 4
pt_mta_ridership_data <- mta_data_long %>%
  filter(Metric_Type == "Total Estimated Ridership")

ggplot(pt_mta_ridership_data %>% filter(Transport_Mode %in% c("Subways", "Buses")), 
       aes(x = Date, y = Value, color = Transport_Mode)) +
  geom_point() +
  labs(title = "Ridership Trends Over Time: Subways vs Buses",
       x = "Date",
       y = "Total Estimated Ridership",
       color = "Transport Mode") 

# Calculate the average ridership for each transport mode during the pandemic
avg_ridership_pandemic <- pt_mta_ridership_data %>%
  filter(Date >= as.Date("2020-03-01") & Date <= as.Date("2023-05-11")) %>%
  group_by(Transport_Mode) %>%
  summarize(Average_Ridership = mean(Value, na.rm = TRUE))

print(avg_ridership_pandemic)

# Analysis 5
pre_pandemic <- pt_mta_ridership_data %>%
  filter(Date >= as.Date("2020-03-01") & Date <= as.Date("2020-03-11"))
print(pre_pandemic)

post_pandemic <- pt_mta_ridership_data %>%
  filter(Date >= as.Date("2023-05-12") & Date <= as.Date("2024-10-10"))
print(post_pandemic)

avg_pre_pandemic <- pre_pandemic %>%
  group_by(Transport_Mode) %>%
  summarize(Average_Ridership_Pre = mean(Value, na.rm = TRUE))
print(avg_pre_pandemic)

avg_post_pandemic <- post_pandemic %>%
  group_by(Transport_Mode) %>%
  summarize(Average_Ridership_Post = mean(Value, na.rm = TRUE))
print(avg_post_pandemic)

avg_ridership <- merge(avg_pre_pandemic, avg_post_pandemic, by = "Transport_Mode")
print(avg_ridership)

avg_ridership <- avg_ridership %>%
  mutate(Percentage_Change = ((Average_Ridership_Post - Average_Ridership_Pre) / Average_Ridership_Pre) * 100)
print(avg_ridership)

ggplot(avg_ridership, aes(x = Transport_Mode, y = Percentage_Change, fill = Transport_Mode)) +
  geom_bar(stat = "identity") +
  labs(title = "Percentage Change in Ridership Due to COVID-19",
       x = "Transport Mode",
       y = "Percentage Change in Ridership")
```

#### Analysis:

##### 1- Analyze the average ridership for each transport mode.

The data shows that subways have the most riders, with about 2.48
million people using them daily, much more than other types of
transport. Buses come next, with around 1 million riders each day,
making them very important. LIRR and Metro-North have fewer riders, with
around 134,000 and 113,000 daily, since they serve commuters in specific
regions. Staten Island Railway has the fewest riders, just over 4,000,
likely because it covers a smaller area. Overall, subways and buses are
the main ways people get around in the city.

##### 2- Analyze how ridership for different transportation modes changes over time. This can help identify patterns, such as the impact of the COVID-19 pandemic on public transportation usage.

As we can see, the chart shows public transportation ridership trends
from 2020 to 2024, highlighting a sharp drop across all modes in early
2020 due to the COVID-19 pandemic, with subway ridership (in purple)
experiencing the most significant decline. Ridership began recovering
mid-2020, with buses showing a steadier recovery compared to the more
volatile subway data. Regional transport modes like LIRR, Metro-North,
and Staten Island Railway have consistently lower ridership. The chart
reveals that, despite gradual recovery, ridership across all modes has
not fully returned to pre-pandemic levels by 2024.

##### 3- Analyze how the percentage of ridership compared to pre-pandemic levels changes over time for each transportation mode.

The chart shows that public transportation usage dropped sharply during
the pandemic but has been recovering at different rates across transport
modes from 2020 to 2024. Bridges and tunnels saw the fastest recovery,
exceeding 100% of pre-pandemic levels by 2021, indicating a shift
towards car travel. Access-A-Ride and buses gradually returned to
normal, nearing or slightly surpassing pre-pandemic levels by 2024.
However, commuter services like the LIRR, Metro-North, and Staten Island
Railway have been slower to recover, remaining below 100%, likely due to
changes in work patterns. Subways are also recovering slowly, still
below pre-pandemic levels by 2024.

##### 4- Analyze the ridership trends between different transportation modes (subways vs buses) to see which modes were more resilient during the pandemic.

The graph shows that both subway and bus ridership dropped sharply at
the start of 2020 due to the pandemic. Subways saw a bigger drop than
buses, but they have been recovering faster. By 2024, subway ridership
has risen back to over 2 million, though it fluctuates more, while bus
ridership has stayed steadier but remains below 2 million. Overall,
subways have more riders than buses, but buses have a more stable number
of users over time.

##### 5- Analyze the impact of the COVID-19 pandemic on ridership by comparing pre-pandemic and post-pandemic data.

The graph illustrates that most MTA transportation systems experienced a
decline in ridership due to the COVID-19 pandemic, with Staten Island
Railroad suffering the largest decrease, with over 50% fewer passengers.
In contrast, Metro-North has seen an increase in ridership, likely due
to people relocating from New York City during the pandemic and opting
to commute using Metro-North.

## Dataset 3: K-12 Schools diversity from 1994-2017 in all states

```{r}
url3<- "https://raw.githubusercontent.com/stormwhale/data-mines/refs/heads/main/school%20divers.csv"
df3<-read.csv(url3)
head(df3)
```

Cleaning and tidying the dataset. The racial groups are already
represented in percentage for each school. NA values are assumed to be
0. All percentage will be rounded to the nearest tenth value.

```{r}
df3_tidy<- df3 %>%
  mutate_all(~replace(., is.na(.), 0)) %>%
  pivot_longer(cols = AIAN:Multi,
               names_to = 'Ethnicity',
               values_to = 'Student_percentage') %>%
  mutate(Student_percentage=round(as.numeric(Student_percentage),1))

head(df3_tidy)
```

### Data analysis:

### 1) Overall ethnicity distribution of K-12 students:

Asian and AIAN groups seem to be the most under integrated among
different schools. (See plot below)

```{r}
ggplot(df3_tidy, aes(x=Student_percentage, fill=Ethnicity)) +
  geom_histogram(binwidth = 5, position='dodge') +
  labs(y='frequency',
       title = 'Frequency distribution of K-12 students ethnicity',
       subtitle = 'Counted by student percentage')
```

### In 2016-2017, which states have schools that integrated at least 50% Asian and AIAN students?

```{r}
df3_2017<- df3_tidy %>%
  filter(SCHOOL_YEAR=='2016-2017') %>%
  filter((Ethnicity =='Asian' | Ethnicity == 'AIAN') & Student_percentage >= 50) %>%
  group_by(ST, Ethnicity) %>%
  summarize(mean_percentage = mean(Student_percentage), .groups = 'drop') %>%
  arrange(desc(mean_percentage))

print(df3_2017)
```

```{r}
ggplot(df3_2017, aes(x=reorder(ST, -mean_percentage), y=mean_percentage, fill= Ethnicity)) +
  geom_bar(stat='identity', position = 'dodge') +
  labs(title = 'In year 2016-2017, states that have schools with at least 50% AIAN or Asian',
       x='States',
       y='Mean percentage of students in each state')
```

### 2) - Between State Comparisons, compare the integration levels of different ethnic groups between states to see which states are more integrated.

```{r between-states}
df3_state_comparison <- df3_tidy %>%
  group_by(ST, Ethnicity) %>%
  summarize(mean_percentage = mean(Student_percentage), .groups = 'drop') %>%
  arrange(desc(mean_percentage))

ggplot(df3_state_comparison, aes(x=reorder(ST, -mean_percentage), y=mean_percentage, fill=Ethnicity)) +
  geom_bar(stat='identity', position='dodge') +
  labs(title = 'Comparison of Ethnic Integration Between States',
       x='States',
       y='Mean Percentage of Students') +
  coord_flip() + 
  theme(axis.text.y= element_text(size = 7))
```

#### Analysis 1: Overall distribution of ethnicity

The graph depicts the mean percentage of students in U.S. states during
the 2016-2017 school year, with at least 50% of either American Indian
and Alaska Native or Asian students integrated into schools. Wyoming,
North Dakota, and Nebraska have the highest proportions of AIAN
students, approaching or at 100%. In contrast, states like Arizona,
California, and New Jersey show substantial integration of Asian
students. States like South Dakota, Washington, and Idaho present a more
balanced distribution between AIAN and Asian student populations.
Overall, the chart highlights significant regional differences in ethnic
integration within school systems, particularly for AIAN and Asian
students

#### Analysis 2: Between State Comparison

The graph compares the ethnic integration of students in various states,
showing the mean percentage of different racial groups in schools. White
students make up the largest proportion of students in most states,
often exceeding 50% and even approaching 100% in several. Hispanic,
Black, and Asian students are represented at lower percentages across
states, with some variation. AIAN students and students of multiple
races have smaller but noticeable percentages in some states. The chart
highlights significant racial differences in student populations across
states, with White students being the most dominant group.
