---
title: "MBA assignment"
author: "Chi Hang(Philip) Cheung"
date: "2025-04-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(arules)
library(arulesViz)
```

```{r}
df<- read.transactions('GroceryDataSet.csv', format = 'basket', sep = ',')
#need to use the inspect() command to view the results in arules:
inspect(head(df,5))
```
## Viewing the top 20 most frequently purchased items:
```{r}
itemFrequencyPlot(df, topN=10, type='absolute', horiz=TRUE) #type = 'relative' will change it to percentage rather than actual numbers
```
#This is to search or list all item frequency distribution:
```{r}
itemFrequency(df)

#You can also search a specific item:
itemFrequency(df)['whole milk']
```
#To search for similar words:
```{r}
itemFrequency(df)[grep('soup', itemLabels(df))]
```

#To mining the rules:
supp = the support--defined as minimal % of carts must contain the all items in the rule. (Or how frequently the items must appear in the carts)
conf = how confident that item B appears along side with item A or with a set of items
minlen = minimal number of items in a rule (usual 2-4 is good)
maxlen = maximum number of items in a rule
```{r}
rules<-apriori(df, parameter = list(supp = 0.001, conf = 0.8, minlen = 2, maxlen = 4), control = list(verbose=FALSE))
```


#To inspect the rules found:
counts = number of carts that satisfy the rule
support = fraction of basket that has the items in the rule. eg. {liquor, red/blush wine} then {bottled beer}
confidence = how confident is this rule? (probility of this item times lift = confidence)
coverage = fraction of the items on the left hand side(basket)
lift = the how likely is the 'tag-along' item appear to be purchased together. eg. lift = 11.2 means bottled beer is 11.2 times more likely to be bought together with liquor, red/blush wine
```{r}
inspect(rules)
```

#Sorting by different criteria:
```{r}
inspect(sort(rules, by = 'lift', decreasing = TRUE)[1:5])
inspect(sort(rules, by = 'count', decreasing = TRUE)[1:5])
```
to visualize:
```{r}
plot(rules, method='graph', engine='htmlwidge')
```

```{r}

```

