---
title: "MBA assignment"
author: "Chi Hang(Philip) Cheung"
date: "2025-04-25"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(arules)
library(arulesViz)
library(cluster)
library(factoextra)
library(tidyverse)
library(ggplot2)
library(tidytext)
```

```{r}
url<- 'https://raw.githubusercontent.com/stormwhale/data-mines/refs/heads/main/GroceryDataSet.csv'
df<- read.transactions(url, format = 'basket', sep = ',')
#need to use the inspect() command to view the results in arules:
inspect(head(df,5))
```

To mining the rules:

supp = the support--defined as minimal % of carts must contain the all
items in the rule. (Or how frequently the items must appear in the
carts). We use 0.1%.

conf = how confident that item B appears along side with item A or with
a set of items. We use 80% confidence.

minlen = minimal number of items in a rule (usual 2-4 is good)

maxlen = maximum number of items in a rule

```{r}
rules<-apriori(df, parameter = list(supp = 0.001, conf = 0.8, minlen = 2, maxlen = 4), control = list(verbose=FALSE))
```

To inspect the top 10 rules based on lift: liquor,red/blush wine has the
strongest link to be purchased with bottled beer. Lift = 11.2

```{r}
inspect(sort(rules, by = 'lift', decreasing = TRUE)[1:10])

```

To visualize:

```{r}
plot(rules[1:10], method='graph', engine='htmlwidge')
```

Cluster analysis:

```{r}
#convert transaction object into a matrix:
trans_matrix<- as(df, 'matrix')
```

Compute Jaccard distance

```{r}
dissimilarity <- dist(trans_matrix, method = "binary")
```

Compute the distances to hierarchical method:

```{r}
hc <- hclust(dissimilarity, method='ward.D2')

#plot the dendrogram:
plot(hc, main = "Dendrogram of Grocery Transactions", xlab = '', sub = '')
```

Dividing the items into 4 clusters and count them: Cluster 1 has the
biggest number of items while 2 to 4 are about the same.

```{r}
#Cut the clusters into 6:
clusters <- cutree(hc, k=4)

table(clusters)
```

Analyze clusters:

```{r}
# Add cluster labels to matrix
clustered_data<- as.data.frame(trans_matrix)
clustered_data$Cluster <- clusters

# View average presence of items per cluster
list1<- aggregate(. ~ Cluster, data = clustered_data, mean)
```

To visualize with the top ten items in the clusters:

```{r}
long_list1<- list1 %>% 
  pivot_longer(cols = -1,
               names_to = 'items',
               values_to = 'average_presence') %>% 
  group_by(Cluster) %>% 
  slice_max(average_presence, n = 10) %>% 
  mutate(items = reorder_within(items, average_presence, Cluster))

long_list1 %>% ggplot(aes(y=items, x=average_presence, fill=as.factor(Cluster)))+
  geom_bar(stat = 'identity')+
  facet_wrap(~Cluster, scales='free_y')+
  scale_y_reordered(labels = function(x) gsub("___\\d+", "", x))+
  labs(title='Top 10 most frequently appear items in four clusters')
  

```

visualize:

```{r}
fviz_cluster(list(data= trans_matrix, cluster = clusters))
```
